### 2015-07-23 Harris_2015/manuscript-materials/draft.tex
TeX

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\textbf{\emph{Abstract}}

Inferring species interactions from observational data is one of the
most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and
produce surprising indirect consequences. For example, two competing
species would ordinarily correlate negatively in space, but this effect
can be reversed in the presence of a third species that is capable of
outcompeting both of them when it is present. Here, I apply models from
statistical physics, called Markov networks or Markov random fields,
that can predict the direct and indirect consequences of any possible
species interaction matrix. Interactions in these models can be
estimated from observational data via maximum likelihood. Using
simulated landscapes with known pairwise interaction strengths, I
evaluated Markov networks and several existing approaches. The Markov
networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect
interactions or abiotic environmental effects largely overpowered them.
A linear approximation, based on partial covariances, also performed
well as long as the number of sampled locations exceeded the number of
species in the data. Indirect effects reliably caused a common null
modeling approach to produce incorrect inferences, however.

@@

\textbf{\emph{Introduction}}

If nontrophic species interactions, such as competition, are important
drivers of community assembly, then ecologists might expect to see their
influence in our data sets (MacArthur 1958, Diamond 1975). Despite
decades of work and several major controversies, however (Lewin 1983,
Strong et al. 1984, Gotelli and Entsminger 2003, Connor et al. 2013),
existing methods for detecting competition's effects on community
structure are unreliable (Gotelli and Ulrich 2009), and thus important
ecological processes remain poorly understood. More generally, it can be
difficult to reason about the complex web of direct and indirect species
interactions in real assemblages, especially when these interactions
occur against a background of other ecological processes such as
dispersal limitation and environmental filtering (Connor et al. 2013).
For this reason, it isn't always clear what kinds of patterns would even
constitute evidence of competition, as opposed to some other biological
process or random sampling error (Lewin 1983, Roughgarden 1983).

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur, versus the frequency that
would be expected if \emph{all} species on the landscape were
independent (Strong et al. 1984, Gotelli and Ulrich 2009). Examining a
species pair against such a ``null'' background, however, rules out the
possibility that the overall association between two species could be
driven by an outside force. For example, even though the two shrub
species in Figure 1 compete with one another for resources at a
mechanistic level, they end up clustering together on the landscape
because they both grow best in areas that are not overshadowed by trees.
If this sort of effect is common, then significant deviations from
independence will not generally provide convincing evidence of species'
direct effects on one another.

While competition between the two shrubs in the previous example does
not leave the commonly-expected pattern in community structure (negative
association at the landscape level), it nevertheless does leave a signal
in the data (Figure 1C). Specifically, \emph{among shaded sites}, there
will be a deficit of co-occurrences, and \emph{among unshaded sites},
there will also be such a deficit.

In this paper, I introduce Markov networks (undirected graphical models
also known as Markov random fields; Murphy 2012) as a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them with observational data. Markov
networks, which generalize partial correlations to non-Gaussian data
(Lee and Hastie 2012, Loh and Wainwright 2013), have been used in many
scientific fields for decades to model associations between various
kinds of ``particles''. For example, a well-studied network named the
Ising model has played an important role in our understanding of physics
(where nearby particles tend to align magnetically with one another;
Cipra 1987). In spatial contexts, these models have been used to
describe interactions between adjacent grid cells (Harris 1974, Gelfand
et al. 2005). In neurobiology, they have helped researchers determine
which neurons are connected to one another by modeling the structure in
their firing patterns (Schneidman et al. 2006). Following recent work by
Azaele et al. (2010) and Fort (2013), I suggest that ecologists could
similarly treat species as the interacting particles in the same
modeling framework. Doing so would allow ecologists to simulate and
study the landscape-level consequences of arbitrary species interaction
matrices, even when our observations are not Gaussian. While ecologists
explored some related approaches in the 1980's (Whittam and
Siegel-Causey 1981), computational limitations had previously imposed
severe approximations that produced unintelligible results (e.g.
``probabilities'' greater than one; Gilpin and Diamond 1982). Now that
it is computationally feasible to fit these models exactly, the approach
has become worth a second look.

@@

\textbf{\emph{Methods}}

\textbf{\emph{Conditional relationships and Markov networks.}}
Ecologists are often interested in inferring direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate
\emph{conditional} (``all-else-equal'') relationships, rather than
\emph{marginal} (``overall'') relationships. The most familiar
conditional relationship is the partial correlation, which indicates the
portion of the sample correlation between two species that remains after
controlling for other variables in the data set (Albrecht and Gotelli
2001). The example with the shrubs and trees in Figure 1 shows how the
two correlation measures can have opposite signs, and suggests that the
partial correlation is more relevant for drawing inferences about
species interactions (e.g.~competition). Markov networks extend this
approach to non-Gaussian data, much as generalized linear models do for
linear regression (Lee and Hastie 2012).

Markov networks give a probability value for every possible combination
of presences and absences in communities. For example, given a network
with binary outcomes (i.e.~0 for absence and 1 for presence), the
relative probability of observing a given presence-absence vector,
\(\vec{y}\), is given by

\[p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).\]

Here, \(\alpha_{i}\) is an intercept term determining the amount that
the presence of species \(i\) contributes to the log-probability of
\(\vec{y}\); it directly controls the prevalence of species \(i\).
Similarly, \(\beta_{ij}\) is the amount that the co-occurrence of
species \(i\) and species \(j\) contributes to the log-probability; it
controls the probability that the two species will be found together
(Figure 2A, Figure 2B). \(\beta\) thus acts as an analog of the partial
covariance, but for non-Gaussian networks. Because the relative
probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends to produce
assemblages that have many pairs of positively-associated species and
relatively few pairs of negatively-associated species (exactly as an
ecologist might expect).

A major benefit of Markov networks is the fact that the conditional
relationships between species can be read directly off the matrix of
\(\beta\) coefficients (Murphy 2012). For example, if the coefficient
linking two mutualist species is \(+2\), then---all else equal---the
odds of observing either species increase by a factor of \(e^2\) when
its partner is present (Murphy 2012). Of course, if all else is
\emph{not} equal (e.g.~Figure 1, where the presence of one competitor is
associated with release from another competitor), then species' marginal
association rates can differ from this expectation. For this reason, it
is important to consider how coefficients' effects propagate through the
network, as discussed below.

Estimating the marginal relationships predicted by a Markov network is
more difficult than estimating conditional relationships, because doing
so requires absolute probability estimates. Turning the relative
probability given by Equation 1 into an absolute probability entails
scaling by a \emph{partition function}, \(Z(\alpha, \beta)\), which is
defined so that that the probabilities of all possible assemblages that
could be produced by the model sum to one (bottom of Figure 2B).
Calculating \(Z(\alpha, \beta)\) exactly, as is done in this paper,
quickly becomes infeasible as the number of species increases: with
\(2^N\) possible assemblages of \(N\) species, the number of bookkeeping
operations required for exact inference spirals exponentially into the
billions and beyond. Numerous techniques are available for working with
Markov networks that keep the computations tractable, e.g.~via analytic
approximations (Lee and Hastie 2012) or Monte Carlo sampling
(Salakhutdinov 2008), but they are beyond the scope of this paper.

\textbf{\emph{Simulations.}} In order to compare different methods for
drawing inferences from observational data, I simulated two sets of
landscapes using known parameters.

The first set of simulated landscapes included the three competing
species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling exactly from a probability
distribution defined by the interaction coefficients in that figure
(Appendix A). Each of the methods described below (a Markov network, two
correlation-based methods and a null model) was then evaluated on its
ability to correctly infer that the two shrub species competed with one
another, despite their frequent co-occurrence.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500,
or 2500 observed communities (24 replicate simulations for each
combination; Appendix B). These simulated data sets span the range from
small, single-observer data sets to large collaborative efforts such as
the North American Breeding Bird Survey. As described in Appendix B, I
randomly drew the ``true'' coefficient values for each replicate so that
most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was three times more common than facilitation.
I then used Gibbs sampling to randomly generate replicate landscapes
with varying numbers of species and sites (Appendix B). For half of the
simulated landscapes, I treated each species' \(\alpha\) coefficient as
a constant, as described above. For the other half, I treated the
\(\alpha\) coefficients as linear functions of two abiotic environmental
factors that varied from location to location across the landscape
(Appendix B). The latter set of simulated landscapes provide an
important test of the methods' ability to distinguish co-occurrence
patterns that were generated from pairwise biotic interactions from
those that were generated by external forces like abiotic environmental
filtering. This task was made especially difficult because---as with
most analyses of presence-absence data for co-occurrence patterns---the
inference procedure did not have access to any information about the
environmental or spatial variables that helped shape the landscape (cf
Connor et al. 2013, Blois et al. 2014).

\textbf{\emph{Inferring \(\alpha\) and \(\beta\) coefficients from
presence-absence data.}} In the previous two sections, the values of
\(\alpha\) and \(\beta\) were known. In practice, however, ecologists
will often need to estimate these parameters from co-occurrence data.
When the number of species is reasonably small, one can compute exact
maximum likelihood estimates for all of the \(\alpha\) and \(\beta\)
coefficients by optimizing \(p(\vec{y}; \alpha, \beta)\). Fully-observed
Markov networks like the ones considered here have unimodal likelihood
surfaces (Murphy 2012), ensuring that this procedure will always
converge on the global maximum. This maximum is the unique combination
of \(\alpha\) and \(\beta\) coefficients that would be expected to
produce exactly the observed co-occurrence frequencies. For the analyses
in this paper, I used the \emph{rosalia} package (Harris 2015a) for the
R programming language (R Core Team 2015) to define the objective
function and gradient as R code. The rosalia package then uses the
\texttt{BFGS} method in R's \texttt{optim} function to find the best
values for \(\alpha\) and \(\beta\).

For analyses with 5 or more species, I made a small modification to the
maximum likelihood procedure described above. Given the large number of
parameters associated with some of the networks to be estimated, I
regularized the likelihood using a logistic prior distribution (Gelman
et al. 2008) with a scale of 1 on the \(\alpha\) and \(\beta\) terms.

\textbf{\emph{Other inference techniques for comparison.}} After fitting
Markov networks to the simulated landscapes described above, I used
several other techniques for inferring the sign and strength of marginal
associations between pairs of species (Appendix B).

The first two alternative interaction measures were the sample
covariances and the partial covariances between each pair of species'
data vectors on the landscape (Albrecht and Gotelli 2001). Because
partial covariances are undefined for landscapes with
perfectly-correlated species pairs, I used a regularized estimate based
on ridge regression {[}Wieringen and Peeters (2014); i.e.~linear
regression with a Gaussian prior{]}. For these analyses, I set the ridge
parameter to 0.2 divided by the number of sites on the landscape.

The third alternative method, described in Gotelli and Ulrich (2009),
involved simulating possible landscapes from a null model that retains
the row and column sums of the original matrix (Strong et al. 1984).
Using the default options in the Pairs software described in Gotelli and
Ulrich (2009), I simulated the null distribution of scaled C-scores (a
test statistic describing the number of \emph{non}-co-occurrences
between two species). The software then calculated a \(Z\) statistic for
each species pair using this null distribution. After multiplying this
statistic by \(-1\) so that positive values corresponded to facilitation
and negative values corresponded to competition, I used it as another
estimate of species interactions.

\textbf{\emph{Method evaluation.}} For the simulated landscapes based on
Figure 1, method evaluation was fairly qualitative: any method whose
test statistic for the two shrubs indicated a negative relationship
passed; other methods failed.

For the larger landscapes, I rescaled the four methods' estimates using
linear regression through the origin so that they all had a consistent
interpretation. In each method, I regressed the ``true'' \(\beta\)
coefficient for each species pair against the model's estimate,
re-weighting the pairs so that each landscape contributed equally to the
rescaled estimate. These regressions yielded squared errors for each of
the 23,140 simulated species pairs, across all conditions and
replicates. I then partitioned these errors into groups defined by the
properties of the simulated landscapes (i.e.~species richness, number of
observed communities, and the presence/absence of environmental
filtering). Within each partition, the mean squared error was used to
calculate the proportion of variance explained (compared with a baseline
model that assumed all interaction strengths to be zero).

\textbf{\emph{Results}}

\textbf{\emph{Three species.}} As shown in Figure 1, the marginal
relationship between the two shrub species was positive---despite their
competition for space at a mechanistic level---due to indirect effects
of the dominant tree species. As a result, the covariance method falsely
reported positive associations for 94\% of the simulated landscapes, and
the randomization-based null model falsely reported such associations
100\% of the time. The two methods for evaluating conditional
relationships (Markov networks and partial covariances), however,
successfully controlled for the indirect pathway via the tree species
and each correctly identified the direct negative interaction between
the shrubs 94\% of the time.

\textbf{\emph{Larger landscapes.}} The accuracy of the four evaluated
methods varied substantially, depending on the parameters that produced
the simulated communities (Figure 3). In general, however, there was a
consistent ordering: overall, the Markov network explained 54\% of the
``true'' parameters' squared deviations from zero, followed by partial
covariances (33\%), and sample covariances (22\%). The null model scores
initially explained only 12\%. After manually reducing the value of one
especially strong outlier (\(Z=1004\), implying \(p<10^{-1000000}\)),
this increased to 17\% (Appendix B). Figure 3 reflects the adjusted
version of the results.

\textbf{\emph{Discussion}}

The results presented above show that Markov networks can reliably
recover species' pairwise interactions from observational data, even for
cases where a common null modeling technique reliably fails.
Specifically, Markov networks were successful even when direct
interactions were largely overwhelmed by indirect effects (Figure 1) or
environmental effects (lower panels of Figure 3). For cases where
fitting a Markov network is computationally infeasible, these results
also indicate that partial covariances---which can be computed
straightforwardly by linear regression---can often provide a
surprisingly useful approximation. The partial correlations' success on
simulated data may not carry over to real data sets, however; Loh and
Wainwright (2013) show that the linear approximations can be less
reliable in cases where the true interaction matrix contains more
structure (e.g.~guilds or trophic levels). On the other hand, if
ecologists are familiar enough with the natural history of their study
systems to describe this kind of structure as a prior distribution on
the parameters or as a penalty on the likelihood, then this information
could reduce the effective degres of freedom to estimate and real-world
results might be even better than those shown in Figure 3.

Ecologists will also need natural history to pin down the exact nature
of the interactions identified by a network model (e.g.~which species in
a positively-associated pair is facilitating the other), particularly
when real pairs of species can reciprocally influence one another in
multiple ways simultaneously (Bruno et al. 2003); the \(\beta\)
coefficients in Markov networks have to reduce this complexity to a
single number. In short, partial correlations and Markov networks both
help prevent us from mistaking marginal associations for conditional
ones, but they can't tell us the underlying biological mechanisms at
work.

Despite these limitations, Markov networks have enormous potential to
improve ecological inferences. For example, Markov networks provide a
simple answer to the question of how competition should affect a
species' overall prevalence, which was a major flash point for the null
model debates in the 1980's (Roughgarden 1983, Strong et al. 1984).
Equation 1 can be used to calculate the expected prevalence of a species
in the absence of biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee
and Hastie 2012). Competition's effect on prevalence in a Markov network
can then be calculated by subtracting this value and the observed
prevalence (cf Figure 2D).

>>>>

\documentclass[12pt,]{article}
\usepackage[compact]{titlesec}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\textbf{Abstract:} Estimating species interactions from observational
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, two competing species would ordinarily correlate negatively in
space, but this effect can be reversed in the presence of a third
species that is capable of outcompeting both of them when it is present.
Here, I apply models from statistical physics, called Markov networks or
Markov random fields, that can predict the direct and indirect
consequences of any possible species interaction matrix. Interactions in
these models can be estimated from observational data via maximum
likelihood. Using simulated landscapes with known pairwise interaction
strengths, I evaluated Markov networks and several existing approaches.
The Markov networks consistently outperformed other methods, correctly
isolating direct interactions between species pairs even when indirect
interactions or abiotic environmental effects largely overpowered them.
Two computationally efficient approximations, based on linear and
generalized linear models, also performed well. Indirect effects
reliably caused a common null modeling approach to produce incorrect
inferences, however.

@@

\setlength{\parindent}{1cm} \setlength{\parskip}{6pt}

\subsubsection{Introduction}\label{introduction}

If nontrophic species interactions, such as competition, are important
drivers of community assembly, then ecologists might expect to see their
influence in our data sets (MacArthur 1958, Diamond 1975). Despite
decades of work and several major controversies, however (Lewin 1983,
Strong et al. 1984, Gotelli and Entsminger 2003, Connor et al. 2013),
existing methods for detecting competition's effects on community
structure are unreliable (Gotelli and Ulrich 2009). In particular,
species' effects on one another can become lost in the complex web of
direct and indirect interactions in real assemblages. For example, the
competitive interaction between the two shrub species in Figure 1A can
become obscured by their shared tendency to occur in unshaded areas
(Figure 1B). If this sort of effect is common, then the vast majority of
ecologists' methods, which rely on test statistics describing the total
number of co-occurrences or overall correlation between putatively
interacting species (Diamond 1975, Strong et al. 1984, Gotelli and
Ulrich 2009, Veech 2013, Pollock et al. 2014) will not generally provide
much evidence regarding species' direct effects on one another.

While competition doesn't reliably leave the expected pattern in
community structure (a low co-occurrence rate at the landscape level),
it nevertheless does leave a signal in the data (Figure 1C).
Specifically, \emph{among shaded sites}, there will be a deficit of
co-occurrences between the competing shrub species, and \emph{among
unshaded sites}, there will also be such a deficit. More generally, we
can obtain much better estimates of the association between two species
from their conditional relationships (i.e.~by controlling for other
species in the network) than we could get from their overall
co-occurrence rates. This kind of precision is difficult to obtain from
null models, which begin with the assumptions that all the pairwise
interactions are zero and thus don't need to be controlled for.
Nevertheless, null models have dominated this field for more than three
decades (Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009).

In this paper, following recent work by Azaele et al. (2010) and Fort
(2013), I show that Markov networks (undirected graphical models also
known as Markov random fields; Murphy 2012) can provide a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observed presence-absence
matrices. Markov networks have been used in many scientific fields for
decades in similar contexts, from physics (where nearby particles
interact magnetically with one another; Cipra 1987) to spatial
statistics (where adjacent grid cells interact with one another; Harris
1974, Gelfand et al. 2005). While community ecologists explored some
related approaches in the 1980's (Whittam and Siegel-Causey 1981),
computational limitations had previously imposed severe approximations
that produced unintelligible results (e.g. ``probabilities'' greater
than one; Gilpin and Diamond 1982). Now that it is computationally
feasible to fit these models exactly, this kind of approach is worth a
second look.

@@

\subsubsection{Methods}\label{methods}

\paragraph{Conditional relationships and Markov
networks.}\label{conditional-relationships-and-markov-networks.}

Ecologists are often interested in learning about direct interactions
between species, controlling for the indirect influence of other
species. In statistical terms, this implies that ecologists want to
estimate \emph{conditional} (``all-else-equal'') relationships, rather
than \emph{marginal} (``overall'') relationships. The example with the
shrubs and trees in Figure 1 shows how the two measures can have
opposite signs, and suggests that conditional relationships are more
relevant for drawing inferences about species interactions
(e.g.~competition).

Markov networks define species interactions in conditional terms, and
use these conditional relationships to determine whether a given
combination of co-occurring species is consistent with their
interactions. Given a set of binary outcomes for each species (i.e.~0
for absence and 1 for presence), the Markov network defines the relative
probability of observing a given presence-absence vector, \(\vec{y}\),
as

\centering
\(p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).\)
\raggedright

Here, \(\alpha_{i}\) is an intercept term determining the amount that
the presence of species \(i\) contributes to the log-probability of
\(\vec{y}\); it directly controls the prevalence of species \(i\).
Similarly, \(\beta_{ij}\) is the amount that the co-occurrence of
species \(i\) and species \(j\) contributes to the log-probability; it
controls the probability that the two species will be found together
(Figure 2A, Figure 2B). Because the relative probability of a
presence-absence vector increases when positively-associated species
co-occur and decreases when negatively-associated species co-occur, the
model tends to produce assemblages that have many pairs of
positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect).

A major benefit of Markov networks is the fact that the conditional
relationships between species can be read directly off the matrix of
\(\beta\) coefficients (Murphy 2012). For example, if the coefficient
linking two mutualist species is \(+2\), then---all else equal---the
odds of observing either species increase by a factor of \(e^2\) when
its partner is present (Murphy 2012). In this way, the \(\beta\)
coefficients behave very similarly to the coefficients in a logistic
regression model (Lee and Hastie 2012).

Of course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from
this expectation. For this reason, it is important to consider how
coefficients' effects propagate through the network, as discussed below.

Estimating the marginal relationships predicted by a Markov network is
more difficult than estimating conditional relationships, because doing
so requires absolute probability estimates. Turning the relative
probability given by Equation 1 into an absolute probability entails
scaling by a normalizing constant, \(Z\), so that that the probabilities
of all possible assemblages will sum to one (bottom of Figure 2B).
Calculating \(Z\) exactly, as is done in these analyses, quickly becomes
infeasible as the number of species increases: with \(2^N\) possible
assemblages of \(N\) species, the number of bookkeeping operations
required spirals exponentially into the billions and beyond. Numerous
techniques are available for keeping the calculations tractable,
e.g.~via analytic approximations (Lee and Hastie 2012) or Monte Carlo
sampling (Salakhutdinov 2008), but they are beyond the scope of this
paper. Alternatively, as noted below, some common linear and generalized
linear methods can also be used as computationally efficient
approximations to the full network.

\paragraph{\texorpdfstring{Estimating \(\alpha\) and \(\beta\)
coefficients from presence-absence
data.}{Estimating \textbackslash{}alpha and \textbackslash{}beta coefficients from presence-absence data.}}\label{estimating-alpha-and-beta-coefficients-from-presence-absence-data.}

In the previous two sections, the values of \(\alpha\) and \(\beta\)
were known. In practice, however, ecologists will often need to estimate
these parameters from co-occurrence data (i.e.~from a matrix of ones and
zeros indicating which species are present at which sites). When the
number of species is reasonably small, one can compute exact maximum
likelihood estimates for all of the \(\alpha\) and \(\beta\)
coefficients given a presence-absence matrix by optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces (Murphy 2012),
ensuring that this procedure will always converge on the global maximum.
This maximum represents the unique combination of \(\alpha\) and
\(\beta\) coefficients that would be expected to produce exactly the
observed co-occurrence frequencies (i.e.~it matches the sufficient
statistics of the model distribution to the sufficient statistics of the
data). For the analyses in this paper, I used the \emph{rosalia} package
(Harris 2015a) for the R programming language (R Core Team 2015) to
define the objective function and gradient in R code and then find the
optimum.

\paragraph{Simulated landscapes.}\label{simulated-landscapes.}

In order to compare different methods, I simulated two sets of
landscapes using known parameters.

The first set of simulated landscapes included the three competing
species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling exactly from a probability
distribution defined by the interaction coefficients in that figure
(Appendix A). Each of the methods described below was then evaluated on
its ability to correctly infer that the two shrub species competed with
one another, despite their frequent co-occurrence.

I also simulated a second set of landscapes using a stochastic community
model based on generalized Lotka-Volterra dynamics, as described in the
Appendix. In these simulations, each species pair was randomly assigned
to either compete for a portion of the available carrying capacity
(negative interaction) or to act as mutualists (positive interaction).
In these simulations, mutualism operated by mitigating the effects of
intraspecific competition on each partner's death rate. For these
analyses, I simulated landscapes with up to 20 species and 25, 200, or
1600 sites (50 replicates per landscape size; see Appendix).

\paragraph{Recovering species interactions from simulated
data.}\label{recovering-species-interactions-from-simulated-data.}

I used Markov networks and several other techniques for determining the
sign and strength of the associations between pairs of species (Appendix
B).

For the Markov networks, I used the rosalia package (Harris 2015a), as
described above. Given the large number of parameters associated with
some of the networks to be estimated, I regularized the likelihood using
a logistic prior distribution with a scale of 2 on the \(\alpha\) and
\(\beta\) terms. This convex prior distribution has a similar shape to
the Laplace distribution used in LASSO regularization (especially in the
tails), but does not force the estimates to be exactly zero.

I also evaluated five alternative methods from the existing literature
to calibrate the performance of the Markov network as a method for
estimating species interactions. The first two alternative interaction
measures were the sample correlations and the partial correlations
between each pair of species' data vectors on the landscape (Albrecht
and Gotelli 2001, Faisal et al. 2010). Because partial correlations are
undefined for landscapes with perfectly-correlated species pairs, I used
a regularized estimate based on James-Stein shrinkage, as implemented in
the corpcor package's \texttt{pcor.shrink} function with the default
settings ({\textbf{???}}). In the context of non-Gaussian data, the
partial correlation can be thought of as a computationally efficient
approximation to the full Markov network model (Loh and Wainwright
2013).

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network (Lee and Hastie 2012). Following Faisal et al. (2010), I fit
regularized logistic regression models for each species, using the other
species on the landscape as predictors. {[}{[}Observational
presence-absence data does not have enough degrees of freedom to
estimate this many parameters, so I symmetrized the pairwise
correlations by averaging the coefficient for predicting species \(i\)'s
status from species \(j\) and vice versa.{]}{]}

The fourth alternative method, described in Gotelli and Ulrich (2009),
involved simulating possible landscapes from a null model that retains
the row and column sums of the original matrix (Strong et al. 1984).
Using the default options in the Pairs software described in Gotelli and
Ulrich (2009), I simulated the null distribution of scaled C-scores (a
test statistic describing the number of \emph{non}-co-occurrences
between two species). The software then calculated a \(Z\) statistic for
each species pair using this null distribution. After multiplying this
statistic by \(-1\) so that positive values corresponded to facilitation
and negative values corresponded to competition, I used it as another
estimate of species interactions.

The last two methods used the latent correlation matrix estimated by the
BayesComm package ({\textbf{???}}) to evaluate the claim from recent
papers that the correlation coefficients estimated by ``joint species
distribution models'' provide an accurate assessment of species'
pairwise interactions (Pollock et al. 2014, see also Harris 2015b). I
reported both the average correlation between each pair of species'
latent variables across 1000 Monte Carlo samples, as well as the average
\emph{partial} correlation.

\paragraph{Evaluating model
performance.}\label{evaluating-model-performance.}

For the simulated landscapes based on Figure 1, method evaluation was
fairly qualitative: any method whose test statistic for the two shrubs
indicated a negative relationship passed; other methods failed. For
methods that provide confidence intervals or other inferential
statistics (Appendix), I also reported how often they rejected the null
hypothesis of no interaction.

Because the different methods mostly describe species interactions on
different scales (e.g.~correlations versus \(Z\) scores versus
regression coefficients), I used linear regression through the origin to
rescale the different estimates produced by each method so that they had
a consistent interpretation. After rescaling each method's estimates, I
calculated squared errors between the scaled interaction estimates and
``true'' interaction values for each of the {[}{[}N{]}{]} species pairs
across all the simulated data sets. These squared errors determined the
proportion of variance explained for different combinations of method
and landscape size (compared with a baseline model that assumed all
interaction strengths to be zero).

\subsubsection{Results}\label{results}

\paragraph{Three species.}\label{three-species.}

As shown in Figure 1, the marginal relationship between the two shrub
species was positive---despite their competition for space at a
mechanistic level---due to indirect effects of the dominant tree
species. As a result, the covariance method falsely reported positive
associations for 94\% of the simulated landscapes, and the
randomization-based null model falsely reported such associations 100\%
of the time. The methods for evaluating conditional relationships
(Markov networks, GLMs, and both types of partial covariance estimates),
however, successfully controlled for the indirect pathway via the tree
species and each correctly identified the direct negative interaction
between the shrubs 94\% of the time.

\paragraph{Twenty species.}\label{twenty-species.}

The accuracy of the four evaluated methods varied substantially,
depending on the parameters that produced the simulated communities
(Figure 3). In general, however, there was a consistent ordering\ldots{}

The three methods that estimated marginal relationships among species
and did not explicitly control for indirect species interactions (the
null model, the joint species distribution model's correlation
coefficients, and the sample correlation coefficients) differed from one
another in their \(R^2\) values, but their estimates were surprisingly
similar to one another. In particular, the correlation coefficients were
extremely similar to the estimates from the other two methods
(Spearman's rho = X for the joint species distribution model's
correlation matrix and Y for the null model).

\subsubsection{Discussion}\label{discussion}

The results presented above show that Markov networks can reliably
recover species' pairwise interactions from observational data, even for
cases where a common null modeling technique reliably fails.
Specifically, Markov networks were successful even when direct
interactions were largely overwhelmed by indirect effects (Figure 1).
For cases where fitting a Markov network is computationally infeasible,
these results also indicate that partial covariances and generalized
linear models (the two methods that estimated conditional relationships
rather than marginal ones) can both provide useful approximations. The
partial correlations' success on simulated data may not carry over to
real data sets, however; Loh and Wainwright (2013) show that the linear
approximations can be less reliable in cases where the true interaction
matrix contains more structure (e.g.~guilds or trophic levels).
Similarly, the approximation involved in using separate generalized
linear models for each species can occasionally lead to catastrophic
overfitting with small-to-moderate sample sizes (Lee and Hastie 2012).
For these reasons, it will usually be best to fit a Markov network
rather than one of the alternative methods when one's computational
resources allow it.

It's important to note that none of these methods can identify the exact
nature of the pairwise interactions (e.g.~which species in a
positively-associated pair is facilitating the other), particularly when
real pairs of species can reciprocally influence one another in multiple
ways simultaneously (Bruno et al. 2003); with compositional data, there
is only enough information to provide a single number describing each
species pair.\\On the other hand, if ecologists are familiar enough with
the natural history of their study systems that they can augment the
likelihood function with a prior distribution or a suitable penalty,
then this information could reduce the effective degrees of freedom to
estimate and real-world results might be even better than those shown in
Figure 3.

Despite these limitations, Markov networks have enormous potential to
improve ecological understanding. For example, Markov networks provide a
simple answer to the question of how competition should affect a
species' overall prevalence, which was a major flashpoint for the null
model debates in the 1980's (Roughgarden 1983, Strong et al. 1984).
Equation 1 can be used to calculate the expected prevalence of a species
in the absence of biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee
and Hastie 2012). Competition's effect on prevalence in a Markov network
can then be calculated by subtracting this value and the observed
prevalence (cf Figure 2D).
